Generating Random Logic Programs Using Constraint Programming

### 2019-12-31 paper/paper.tex
Counting: It's working!
Renaming

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\mathcal{M}_{\mathcal{A}}}
\newcommand{\maxNumNodes}{\mathcal{M}_{\mathcal{N}}}
\newcommand{\maxNumClauses}{\mathcal{M}_{\mathcal{C}}}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We also use Choco~4.10.2 \cite{choco}. This works with both Prolog
\cite{DBLP:books/daglib/0041598} and ProbLog \cite{DBLP:conf/ijcai/RaedtKT07}.
Tested with SWI-Prolog \cite{DBLP:journals/tplp/WielemakerSTL12}.

@@

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$. All decision
variables of the model are contained in two arrays of length $\maxNumClauses{}$:
\begin{itemize}
\item \variable{Body[] bodiesOfClauses}
\item \variable{Head[] headsOfClauses}
\end{itemize}

@@

\begin{constraint}
  Each predicate gets at least one clause. Let $P = \{ h.\variable{predicate}
  \mid h \in \variable{clauseHeads} \}$. Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

@@

\section{Atoms} % TODO: rename to 'nodes'

%, where $\Box$ means the position is either
%  reserved for a variable, or disabled

\begin{definition} \label{def:atom}
  An \emph{atom} is a $\variable{predicate} \in \tokens{} \cup \predicates{}$
  and a list of \variable{arguments} of length $\maxArity{}$ in $\variables{}
  \cup \constants{}$. The atom's \variable{arity} is a number in $[0,
  \maxArity{}]$ defined by a \variable{table} constraint, according to the
  \variable{predicate}.
\end{definition}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item \variable{treeStructure}: list of length $\maxNumNodes{}$ with
    domain $[0, \maxNumNodes{} - 1]$.
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item \variable{treeValues}: $\maxNumNodes{}$ atoms.
  \end{itemize}
\end{definition}
% TODO: examples

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$, if $\variable{numNodes} \le
  i$, then
  \[
    \variable{treeStructure}[i] = i \quad \text{and} \quad
    \variable{treeValues}[i].\variable{predicate} = \top,
  \]
  else
  \[
    \variable{treeStructure}[i] < \variable{numNodes}.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{itemize}
  \item has 0 children $\iff$ $\variable{treeValues}[i].\variable{predicate} \in
    \predicates{}$;
  \item has 1 child $\iff$ $\variable{treeValues}[i].\variable{predicate} = \neg$;
  \item has $>1$ child $\iff$ $\variable{treeValues}[i].\variable{predicate} \in
    \{ \land, \lor \}$.
  \end{itemize}
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies
    \variable{treeValues}[i].\variable{predicate} \ne \top.
  \]
\end{constraint}

If the clause should be disabled, $\variable{numNodes} = 1$ and
$\variable{treeValues}[0].\variable{predicate} = \top$.

\begin{constraint}
  Adjacency matrix representation:
  \begin{align*}
    A[i][j] = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{predicate} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{constraint}

\section{Head of a Clause}

Our definition of a head of a clause is more restrictive than \cref{def:atom}.

\begin{definition} % TODO: variables are always ordered, constants are not.
  The \emph{head} of a clause is defined by two lists:
  \begin{itemize}
  \item $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause.
  \item \variable{variables} of length $|\variables{}|$ and with domain
    $[0, \maxArity{}]$: how many times each variable appears in the head atom.
  \item \variable{constants} of length $\maxArity{}$ and with domain
    $\constants{} \cup \{ \Box \}$, where $\Box$ denotes that the position is
    reserved for a variable.
  \end{itemize}
  We also define the predicate's \variable{arity} using the same
  \variable{table} constraint.
\end{definition}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{constants}[i] = 0.
  \]
\end{constraint}

\begin{constraint}
  Connecting the two lists:
  \[
    \sum_{v \in \variable{variables}} v = \variable{count}(\Box,
    \variable{constants}) + \variable{arity} - \maxArity{}.
  \]
\end{constraint}

@@

Number of atoms:
\[
  A = \sum_{a \in \arities{}} p_a(|\variables{}| + |\constants{}|)^a
\]

Number of clauses:
\[
  C = 1 + \sum_{n=1}^{\maxNumNodes{}} T(n),
\]
where $T(n)$ is defined recursively as:
\[
  T(1) = A
\]
and
\[
  T(n) = T(n-1) + 2\sum_{\substack{\text{ordered partitions}\\
      c_1 + \dots + c_k = n - 1,\\
      k \ge 2}} \prod_{i=1}^k T(c_i).
\]

@@

Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v=1}^a \binom{a}{v} |\constants{}|^{a-v}
  \sum_{k=0}^{|\variables{}|-1} \binom{v-1}{k}.
\]
First, select the $v$ positions dedicated for variables. The remaining $a-v$
constants can then be filled in $|\constants{}|^{a-v}$ ways. Filling $v$
positions with $k+1$ variables in a non-decreasing manner (without skipping any
variables) can be seen as putting $k$ `bars' in the $v-1$ spaces between $v$
positions. Each bar represents switching to the next variable.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
%      h_1 + \dots + h_{|\predicates{}|} = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \binom{C}{h_i} H_{a_i}^{h_i}
\]

>>>>

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\arrayd}[3]{\variable{{#1}[}{#2}\variable{]} \in {#3}}
\newcommand{\arrayt}[3]{\variable{{#3}[}{#2}\variable{] {#1}}}
\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\mathcal{M}_{\mathcal{A}}}
\newcommand{\maxNumNodes}{\mathcal{M}_{\mathcal{N}}}
\newcommand{\maxNumClauses}{\mathcal{M}_{\mathcal{C}}}

\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$.

We also use Choco~4.10.2 \cite{choco}. This works with both Prolog
\cite{DBLP:books/daglib/0041598} and ProbLog \cite{DBLP:conf/ijcai/RaedtKT07}.
Tested with SWI-Prolog \cite{DBLP:journals/tplp/WielemakerSTL12}.

@@

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$. All decision
variables of the model are contained in two arrays:
\begin{itemize}
\item $\arrayt{bodiesOfClauses}{\maxNumClauses{}}{Body}$,
\item $\arrayt{headsOfClauses}{\maxNumClauses{}}{Head}$
\end{itemize}

@@

\begin{constraint}
  Each predicate gets at least one clause. Let $P = \{ h.\variable{predicate}
  \mid h \in \variable{headsOfClauses} \}$. Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{treeStructure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{treeValues}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}
% TODO: examples

\subsection{Nodes}
%, where $\Box$ means the position is either
%  reserved for a variable, or disabled

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{}}$. The node's
  \variable{arity} is a number in $[0, \maxArity{}]$ defined by a
  \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{name} & \text{if } \variable{name} \in
      \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

@@

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$, if $\variable{numNodes} \le
  i$, then
  \[
    \variable{treeStructure}[i] = i \quad \text{and} \quad
    \variable{treeValues}[i].\variable{name} = \top,
  \]
  else
  \[
    \variable{treeStructure}[i] < \variable{numNodes}.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{itemize}
  \item has 0 children $\iff$ $\variable{treeValues}[i].\variable{name} \in
    \predicates{}$;
  \item has 1 child $\iff$ $\variable{treeValues}[i].\variable{name} = \neg$;
  \item has $>1$ child $\iff$ $\variable{treeValues}[i].\variable{name} \in
    \{ \land, \lor \}$.
  \end{itemize}
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies
    \variable{treeValues}[i].\variable{name} \ne \top.
  \]
\end{constraint}

If the clause should be disabled, $\variable{numNodes} = 1$ and
$\variable{treeValues}[0].\variable{name} = \top$.

\begin{constraint}
  Adjacency matrix representation:
  \begin{align*}
    A[i][j] = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{constraint}

\section{Heads of Clauses}

Our definition of a head of a clause is more restrictive than \cref{def:node}.

\begin{definition} % mention: variables are always ordered, constants are not.
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause.
  \item $\arrayd{variables}{|\variables{}|}{[0, \maxArity{}]}$: how many times
    each variable appears in the head atom.
  \item $\arrayd{constants}{\maxArity{}}{\constants{} \cup \{ \Box \}}$, where
    $\Box$ denotes that the position is reserved for a variable.
  \end{itemize}
  We also define the \variable{predicate}'s \variable{arity} using the same
  \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      0 & \text{if } \variable{predicate} = \Box\\
      \text{the arity of } \variable{predicate} & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{constants}[i] = \Box.
  \]
\end{constraint}

\begin{constraint}
  Connecting the two lists:
  \[
    \maxArity{} - \variable{arity} + \sum_{v \in \variable{variables}} v
    = \begin{cases}
      \maxArity{} & \text{if } \variable{predicate} = \Box \\
      \variable{count}(\Box, \variable{constants}) & \text{otherwise}.
      \end{cases}
  \]
\end{constraint}

@@

Number of atoms (leaves of the tree):
\[
  A = \sum_{a \in \arities{}} p_a(|\variables{}| + |\constants{}|)^a
\]

Number of clauses:
\[
  C = 1 + \sum_{n=1}^{\maxNumNodes{}} T(n),
\]
where $T(n)$ is defined recursively as:
\[
  T(1) = A
\]
and
\[
  T(n) = T(n-1) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      k \ge 2,\\
      c_i \ge 1 \text{ for all } i}} \prod_{i=1}^k T(c_i).
\]

@@

Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = |\constants{}|^a + \sum_{v=1}^a \binom{a}{v} |\constants{}|^{a-v}
  \sum_{k=0}^{|\variables{}|-1} \binom{v-1}{k}.
\]
First, select the $v$ positions dedicated for variables. The remaining $a-v$
constants can then be filled in $|\constants{}|^{a-v}$ ways. Filling $v$
positions with $k+1$ variables in a non-decreasing manner (without skipping any
variables) can be seen as putting $k$ `bars' in the $v-1$ spaces between $v$
positions. Each bar represents switching to the next variable.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
%      h_1 + \dots + h_{|\predicates{}|} = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{C}{h_i} H_{a_i}^{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n + k - 1}{k}.
\]
